---
title: "Basic Analysis for Scale Archiving"

author: 
  - name: Julian Urban
  - name: Piotr Koc
  - name: David Grüning
   
date: 2024-07-01
date-format: "DD.MM.YYYY"
format:
  html:
    code-overflow: wrap
    code-copy: true
    toc: true
    theme: cosmo
    html-math-method: katex
license: "CC BY-NC 4.0"
citation: 
    type: "document"
    title: |
        Basic Analysis for Scale Archiving
    author:
       - name: Julian Urban
       - name: Piotr Koc
       - name: David Grüning
    issued: 2024
    container-title: ZIS R-tutorials
    publisher: GESIS – Leibniz Institute for the Social Sciences 
    URL: https://zis.gesis.org/infotexte/GuidelineMaterials.html
    
---

## Introduction

The following tutorial presents the basic analysis required for an instrument to be archived in the Open Repository for Measurement Instruments. As an example, we use a scale with continuous indicators, so not all parts of the analysis are appropriate for instruments with categorical indicators. We point this out whenever relevant. The treatment of categorical indicators is covered in a separate tutorial. 

## Data Preparation & Descriptive Analysis
First, load the required packages:
```{r, message=FALSE}
  library(lavaan)
  library(psych)
```

In this tutorial, we will use the `HolzingerSwineford1939` data set, which you can access 

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
